Gait Monitoring using Wireless Sensors\\ \vspace{4mm} \large{Final Report 18-748}\\ \large{Spring 2017}

### 2017-05-10 paper.tex
Updated bibliography and makefile

\maketitle

%Only including an abstract since it looks like the submission site wants one... 
\begin{abstract}
In this project we demonstrate a gait monitoring system comprised of Firefly wireless
sensor motes that relay data from inertial measurement units mounted on a subject's body
to a central server. Using the acceleration and orientation information reported at six
different points along a subject's legs and feet, we were able to visualize the subject's
movements in real time and capture the relevant motion data to be used for future
analysis. Our system also provides a low-processing solution for fall detection through
thresholding analysis of the orientation measurements reported by a subset of the sensor
nodes in the system. After testing our system on three different subjects, we believe it
represents a proof of concept of low cost gait monitoring systems that will improve the
quality of patient care in a multitude of settings. 
\end{abstract}

@@

Wireless Body Area Networks have been proposed as a promising solution for capturing
useful health status indicators without inconveniencing a patient~\cite{wban}. The
additional data enables a new field of computer-assisted physical therapy that allows
therapists to deliver more targeted care to their patients. Advanced treatment centers for
patients with unique cases, such as the Walter Reed Center for Performance and Clinical
Research (CPCR) amputee clinic, use extensive computer modeling to aid in patient
rehabilitation~\cite{cpcr}.  An apparent barrier to the widespread adoption of computer
assisted therapy is the cost of the motional analysis systems. For instance, the CPCR uses
23 infrared cameras to track the movement of reflective markers attached to a patient's
body and six force plates in the floor to fully capture a patient's movement. Though the
movement data can be captured and used at a later time, the gait analysis is still limited
to a clinical setting.

Our project's goal was to expand the usefulness of gait analysis beyond a professional’s
office to an individual’s day-to-day environment. We accomplished our goal by
demonstrating a proof-of-concept gait monitoring system that relies on wireless motes
instrumented with Inertial Measurement Units (IMUs) to capture relevant data and relay
them to a powerful backend machine for offline and real time processing. Our system
extracted a variety of components of the subject's gait pattern from linear and angular
acceleration data measured at specfic points along his or her legs. The subject's
movements were then visualized and features of the gait were plotted in real time. We also
demonstrate the practicality of real-time fall detection by using a simple thresholding
technique to indicate when a subject is no longer upright. Throughout the trials with our
prototype system we note that wireless sensor networks are particularly promising for
accomplishing gait detection because the small nodes and lack of constraining wires
minimizes the impact on the patient and allows for an honest gait assessment without the
residual impact of cumbersome measurement devices. 
%
%Add note about final system architecture in Intro... should make the technical
%description easier and it still make sense to do that here
%
\section{Related Work}
This project is based on prior work in gait analysis and wireless sensors networks used to
aid healthcare objectives. In this section we will describe how the prior work influenced
the design of our project 
\subsection{Gait Analysis}
Gait analysis using wearable sensors is a well-studied approach to characterizing gait
patters~\cite{gaits}. The goal of the gait analysis approaches is to analyze the various phases of
the human gait to identify abnormalities. A normal walking gait is comprised of eight
stages as shown in Figure~\ref{fig:phases}.

>>>>

\maketitle
\thispagestyle{plain}
\pagestyle{plain}

%Only including an abstract since it looks like the submission site wants one... 
\begin{abstract}
In this project we demonstrate \sys, a gait monitoring system comprised of Firefly wireless
sensor motes that relay data from inertial measurement units mounted on a subject's body
to a central server. Using the acceleration and orientation information reported at six
different points along a subject's legs and feet, we were able to visualize the subject's
movements in real time and capture the relevant motion data to be used for future
analysis. Our system also provides a low-processing solution for fall detection through
thresholding analysis of the orientation measurements reported by a subset of the sensor
nodes in the system. After testing our system on three different subjects, we believe it
represents a proof of concept of low cost gait monitoring systems that will improve the
quality of patient care in a multitude of settings. 
\end{abstract}

@@

Wireless Body Area Networks have been proposed as a promising solution for capturing
useful health status indicators without inconveniencing a patient~\cite{msban,wban}. The
additional data enables a new field of computer-assisted physical therapy that allows
therapists to deliver more targeted care to their patients. Advanced treatment centers for
patients with unique cases, such as the Walter Reed Center for Performance and Clinical
Research (CPCR) amputee clinic, use extensive computer modeling to aid in patient
rehabilitation~\cite{cpcr}.  An apparent barrier to the widespread adoption of computer
assisted therapy is the cost of the motional analysis systems. For instance, the CPCR uses
23 infrared cameras to track the movement of reflective markers attached to a patient's
body and six force plates in the floor to fully capture a patient's movement. Though the
movement data can be captured and used at a later time, the gait analysis is still limited
to a clinical setting.

\sys's goal was to expand the usefulness of gait analysis beyond a professional’s
office to an individual’s day-to-day environment. We accomplished our goal by
demonstrating a proof-of-concept gait monitoring system that relies on wireless motes
instrumented with Inertial Measurement Units (IMUs) to capture relevant data and relay
them to a powerful backend machine for offline and real time processing. Our system
extracted a variety of components of the subject's gait pattern from linear and angular
acceleration data measured at specfic points along his or her legs. The subject's
movements were then visualized and features of the gait were plotted in real time. We also
demonstrate the practicality of real-time fall detection by using a simple thresholding
technique to indicate when a subject is no longer upright. Throughout the trials with our
prototype system we note that wireless sensor networks are particularly promising for
accomplishing gait detection because the small nodes and lack of constraining wires
minimizes the impact on the patient and allows for an honest gait assessment without the
residual impact of cumbersome measurement devices. 
\section{Related Work}
This project is based on prior work in gait analysis and wireless sensors networks used to
aid healthcare objectives. In this section we will describe how the prior work influenced
the design of our project 
\subsection{Gait Analysis}
Gait analysis using wearable sensors is a well-studied approach to characterizing gait
patters~\cite{gaits}. The goal of the gait analysis approaches is to analyze the various phases of
the human gait to identify abnormalities. A normal walking gait is comprised of eight
stages as shown in Figure~\ref{fig:phases}.
